---
title: "01-Nonstationary time series model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#- Seasonal AR(1) model

**Fit a seasonal AR(1) model – i.e., nonstationary time series model to the monthly Colorado River Flow at Lees Ferry.**


* **Generate 250 simulations each of same length as the historical data.**

```{r results=FALSE}

# Load libraries
libr=c("magrittr","sm","stats","moments")
options(warn=1)
suppressPackageStartupMessages(lapply(libr, require, character.only = TRUE))

```

```{r fig.height=8}
# import and set up flow data

flow = read.table(
  "http://civil.colorado.edu/~balajir/CVEN6833/HWs/HW-3-2018/LeesFerry-monflows-1906-2016.txt")

flow = flow[,2:13] %>% `rownames<-`(flow[,1]) %>%
  setNames(.,c("jan","feb","mar","apr","may","jun",
                      "jul","aug","sep","oct","nov","dec")) %>%
                      {./10^6} # convert AF to MAF
head(flow,n=1L) # show values
tail(flow,n=1L)

flow$year = rowSums(flow) # add year in 13th column

par(mfrow=c(4,3)) # plot histogram and density of monthly flow
for(i in 1:12){
  hist(flow[,i], freq=FALSE,
       main = colnames(flow)[i],xlab = "",ylab = "")
  sm.density(flow[,i], add=TRUE)
  }
```
```{r}
par(mfrow=c(1,1)) # plot histogram and density of anual flow
  hist(flow[,13], freq=FALSE,
              main = colnames(flow)[13],xlab = "")
  sm.density(flow[,13], add=TRUE)

```
```{r}
# Get the parameters of the Thomas Fiering Model (12 models, 1 for each transition)

coef1 = coef2 = rep(0,length.out = 12)

coef1[1] = cor(flow[-1,"jan"],flow[-111,"dec"]) # jan - dec
coef2[1] = sqrt((var(flow[,1])) * (1. - coef1[1]*coef1[1]))

for(i in 2:12){ # all the remaining months
    coef1[i] = cor(flow[,i],flow[,i-1])
    coef2[i] = sqrt((var(flow[,i])) * (1. - coef1[i]*coef1[i]))
}
```

The 12 pairs of coefficients are used to run 250 simulations (synthetic values) that will populate the statistics for the models. 

```{r}

nsim=250	# number of simulations
nyrs=length(flow[,1]) # years 

armean=matrix(0,nsim,12)	#matrices that store the statistics
arstdev=matrix(0,nsim,12)
arcor=matrix(0,nsim,12)
arskw=matrix(0,nsim,12)
armax=matrix(0,nsim,12)
armin=matrix(0,nsim,12)

# Points where May PDF is evaluated 
xeval=seq(min(flow$may)-0.25*sd(flow$may),
          max(flow$may)+0.25*sd(flow$may),length=100)
simpdf=matrix(0,nrow=nsim,ncol=100) # Array to store May simulated PDF


## Simulations

for(k in 1:nsim){ 
  nmons=nyrs*12	#number of values to be generated
  xsim=1:nmons
  r=sample(1:nyrs,1)

  xsim[1]=flow[r,1] # Starting point for sim
  xprev=xsim[1]
  
  for(i in 2:nmons){
    j=i %% 12
    if(j == 0) j=12
    j1=j-1
    
    if(j == 1) j1=12
    
    x1=xprev-mean(flow[,j1])
    x2=coef2[j]*rnorm(1,0,1)
    
    xsim[i]=mean(flow[,j]) + x1*coef1[j] + x2
    xprev=xsim[i]
  }
  
  #Store simulated values in matrix form, get May values and PDF
  simdismon=matrix(xsim,ncol = 12, byrow = TRUE) # filled by row
  maysim = simdismon[,5] # Synthetic values for May
	simpdf[k,]=sm.density(maysim,eval.points=xeval,display="none")$estimate
	
	# Fill statistics for each month
	for(j in 1:12){
	  armean[k,j]=mean(simdismon[,j])
    armax[k,j]=max(simdismon[,j])
    armin[k,j]=min(simdismon[,j])
    arstdev[k,j]=sd(simdismon[,j])
		arskw[k,j]=skewness(simdismon[,j])
	}
	
	arcor[k,1]=cor(simdismon[-nyrs,12],simdismon[2:nyrs,1]) #cor dec-jan
	
	for(j in 2:12){ # rest of pairs
	  j1=j-1
    arcor[k,j]=cor(simdismon[,j],simdismon[,j1])
	}
}
```

To compare the statistics from the synthetic values with the historical data, they are bound together in the same matrix.

```{r}
#Compute statistics from the Historical DAta.
obsmean=1:12
obsstdev=1:12
obscor=1:12
obsskw=1:12
obsmax=1:12
obsmin=1:12

for(i in 1:12){
  obsmax[i]=max(flow[,i])
  obsmin[i]=min(flow[,i])
  obsmean[i]=mean(flow[,i])
  obsstdev[i]=sd(flow[,i])
	obsskw[i]=skewness(flow[,i])
}

for(i in 2:12){
  i1=i-1
  obscor[i]=cor(flow[,i], flow[,i1])
}

obscor[1]= cor(flow[-nyrs,12], flow[2:nyrs,1])

### bind the stats of the observed data at the top..
        armean=rbind(obsmean,armean)
        arstdev=rbind(obsstdev,arstdev)
        arskw=rbind(obsskw,arskw)
        arcor=rbind(obscor,arcor)
        armax=rbind(obsmax,armax)
        armin=rbind(obsmin,armin)

#write the stats in separate files..

write(t(arcor),file="arcorrs1",ncol=12)
write(t(armean),file="armeans",ncol=12)
write(t(arstdev),file="arstdevs",ncol=12)
write(t(arskw),file="arskews",ncol=12)
write(t(armax),file="armax",ncol=12)
write(t(armin),file="armin",ncol=12)
```

* **Create boxplots of annual and monthly, mean, variance, skew, lag-1 correlation, minimum, maximum and PDFs of May and annual flows. Comment on what you observe and also on why some of the monthly statistics are not captured.**

* **For fun, replace the simulation of the errors (or innovations) from Normal to Gamma.**
