---
title: "HW#2e2-main"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```

## Exercise 2. Bayesian Hierarchical Spatial Model
### (Do HW#1.12) Develop a BHSM with a Binomial distribution in the first level

Initialize (source) libraries, functions:
```{r }
setwd("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2")
source("source_file.R")
set.seed(1) # Set seed to repodruce results
```

### Fit a variogram on the residuals of a best GLM
The best GLM (HW#1-8) for binary data was the one with 7 covariates, Binomial family and Cauchit link function (code HW#1 GLM_fit.R: 
Y ~ Lat + Long + Elev + LatLong + LatElev + LongElev + All.

```{r}
# # Adding columns to daily precipitation and fitting best GLM 
# daily$LatLong=daily$Lat*daily$Long
# daily$LatElev=daily$Lat*daily$Elev
# daily$LongElev=daily$Long*daily$Elev
# daily$All=daily$Lat*daily$Long*daily$Elev

# Fit best GLM
yglm<-GLM_fit(daily,"binomial")
summary(yglm)
# Fit a variogram 
geod = as.geodata(cbind(daily$Long,daily$Lat,residuals(yglm,type="response")))
# geod = jitterDupCoords(geod,max=0.01) #jitter close values
vg = variog(geod,uvec = 12)
plot(vg)

```

With the results from the variogram we can assign the initial value of the Monte Carlo Markov Chain
```{r}
# Sigma, phi, (nugget set to zero) 
# sigma.sq = median(vg$v)   # Ask why no sigma with binomial data
# sigma.sq.lo = 0.25*sigma.sq
# sigma.sq.hi = 1.75*sigma.sq
phi.val = median(vg$u) 
phi.lo = 0.25*phi.val
phi.hi = 1.75*phi.val

n.samples <- 500 # Input number of samples
coords <- cbind(daily$Long,daily$Lat) # coordinates of observations

# prior distributions defined (limited distributions available... see help(spLM)
priors <- list("beta.flat", "phi.unif"=c(phi.lo,phi.hi),
               "sigma.sq.ig"=c(100, 100)) 

# starting values for MCMC resampling
starting <- list("phi"=phi.val, "sigma.sq"=500)#, "tau.sq"=100)

# adjustment factor for MCMC sampling routine
tuning <- list("phi"=1, "sigma.sq"=1)#, "tau.sq"=100) # WJR why set these to 1??
```

### Perform Monte Carlo Markov Chain Analysis and plot posterior values

```{r}
require(spBayes)
bat.fm <- spLM(formula=yglm$y~., data=yglm$data, coords=coords, priors=priors, tuning=tuning, starting=starting, cov.model = "exponential", n.samples = n.samples, verbose = FALSE, n.report = 50)
# burn-in samples
bat.fm <- spRecover(bat.fm,start=((1/3)*n.samples)+1, thin=1)

# plot posterior pdfs
beta = bat.fm$p.beta.recover.samples
theta = bat.fm$p.theta.recover.samples
par(mfrow=c(2,2))
plot(beta,auto.layout=FALSE)
plot(theta)
par(mfrow=c(1,1))
```
The results from the MCMC are used to predict values on the grid using a Krigging model on the residuals (This chunk is not evaluated in the knit version and the results are incorporated in the next chunk from a backup file containing the results calculated in the console).

```{r eval=FALSE}
n.samples2 = floor(n.samples*(2/3)) # consider only the last 2/3 of the MCMC simulations

yHBM = matrix(nrow=length(dem[,1]), ncol=n.samples2) # store y values at each dem point for each sample
ySE = matrix(nrow=length(dem[,1]), ncol=n.samples2) # same for SE

xdem=dem[,c("Long","Lat")]
dem$LatLong=dem$Lat*dem$Long #creating DEM with 7 variables as xx
dem$LatElev=dem$Lat*dem$Elev
dem$LongElev=dem$Long*dem$Elev
dem$All=dem$Lat*dem$Long*dem$Elev

# xx=yglm$data # ind.var for Krigging regression model
xx=coords # ind.var for Krigging spatial model
# Loop over Krigging at each sample point
for(i in 1:n.samples2){
  ykrig=Krig(xx,residuals(yglm,type="response"),rho = theta[i,1],theta = theta[i,2],m = 1)    # Krig model
  y2=predict.Krig(ykrig,x=xdem,drop.Z = TRUE,yM=ykrig$yM) # Predict residual from Krig at DEM locations
  ySE[,i]=predictSE(ykrig,x=xdem,drop.Z = TRUE)           # Predict SE of spat Krig at DEM 
  # ySE[,i]=predictSE(ykrig,x=dem,drop.Z = TRUE)           # Predict SE of reg Krig at DEM 

    
  y1=beta[i,1]+dem$Lat*beta[i,2]+dem$Long*beta[i,3]+dem$Elev*beta[i,4]+
                   dem$LatLong*beta[i,5]+dem$LatElev*beta[i,6]+dem$LongElev*beta[i,7]+
                   dem$All*beta[i,8]                           # Predict y from MCMC
  yHBM[,i]=y1+y2                                                  # Add MCMC and residuals
  print(sprintf("Iteration %s complete",i))
      }
```
The results from krigging over the MCMC sample are used to plot the posterior parameters of the Bayesian Hierarchical Spatial Model: histogram of precipitation, spatial map of mean, and spatial map of standard error.

```{r}
load("C:/Users/alexb/Google Drive/CVEN 6833 ADAT/zz Homeworks/HW#2/CVEN6833-HW-2/backup_e2.RData")
yHBM=yHBM #backup files
ySE=ySE
mean.y = apply(yHBM, 1, FUN = mean) # posterior mean of precipitation
mean.yse = apply(ySE, 1, FUN = mean) # posterior standard error

hist(yHBM, main = "Histogram of Precipitation Posterior")

#Create spatial data to plot
X_sp<-getSpatialDataFrame(cbind(xdem,mean.y))
X_sp<-NAD83(X_sp)
X_ppp<-as.ppp(data.frame(X_sp@coords,X_sp@data$`mean.y`), W=colf_owin)

zz_sp<-getSpatialDataFrame(cbind(xdem,mean.yse))
zz_sp<-NAD83(zz_sp)
zz_ppp<-as.ppp(data.frame(zz_sp@coords,zz_sp@data$`mean.yse`), W=colf_owin)

# plot calls
generateplot(X_ppp)
title(main ='BHSM - Predicted Probability of Precip. (as a fraction)')

generateplot(zz_ppp)
title(main='January - BHSM - Predicted Standard Error of Precip. (as a fraction)')
```

