---
title: "02-Non_param_AR1"
output: html_document
---

# Nonparametric seasonal lag-1 model

**Fit a nonparametric seasonal lag-1 model and repeat [exercise] 1. You can use either the the K-nn bootstrap technique or LOCFIT/residual resampling and repeat 1.**

## Generate 250 simulations each of same length as the historical data.

```{r results=FALSE}
# Load libraries
libr=c("magrittr","sm","stats","moments")
options(warn=1)
suppressPackageStartupMessages(lapply(libr, require, character.only = TRUE))
```
```{r}
# import and set up flow data
flow = read.table(
  "http://civil.colorado.edu/~balajir/CVEN6833/HWs/HW-3-2018/LeesFerry-monflows-1906-2016.txt")

flow = flow[,2:13] %>% `rownames<-`(flow[,1]) %>%
  setNames(.,c("jan","feb","mar","apr","may","jun",
                      "jul","aug","sep","oct","nov","dec")) %>%
                      {./10^6} # convert AF to MAF
flow$year = rowSums(flow) # add year in 13th column
head(flow,n=1L) # show values
tail(flow,n=1L)
```
```{r}
# Simulations (innovation defines st. distribution of error)
nsim=100	# number of simulations
nsim1=nsim+1
nyrs=length(flow[,1]) # years 
nyrs1 = nyrs-1
N=nyrs*12 # time series for monthly flow

armean=matrix(0,nsim,12)	#matrices that store the statistics
arstdev=matrix(0,nsim,12)
arcor=matrix(0,nsim,12)
arskw=matrix(0,nsim,12)
armax=matrix(0,nsim,12)
armin=matrix(0,nsim,12)

ar.year.stat=matrix(NA,ncol = 6,nrow = nsim) # year statistics
colnames(ar.year.stat) = c("mean","stdev","min","max","skew","cor")

# Points where May PDF is evaluated 
xeval=seq(min(flow$may)-0.25*sd(flow$may),
          max(flow$may)+0.25*sd(flow$may),length=100)
simpdf=matrix(0,nrow=nsim,ncol=100) # Array to store May simulated PDF

# Points where anual PDF is evaluated 
yeval=seq(min(flow$year)-0.25*sd(flow$year),
          max(flow$year)+0.25*sd(flow$year),length=100)
year.pdf=matrix(0,nrow=nsim,ncol=100) # Array to store anual simulated PDF

# # The anual flow is modeled using single AR(1) model
# ar.year=ar(flow$year,order.max = 1) #AR order 1, MA

# Fitting Np-AR-1 model

K=round(sqrt(nyrs))		#number of nearest neighbors
W= 1:K %>% {1/.} %>% {./sum(.)} %>% cumsum(.) # weight function

for (isim in 1:nsim) {

	i=round(runif(1,2,nyrs)) # initial random year
	
	zsim = 1:N # n of monthly flow time series
	zsim[1] = flow[i,1] # initial flow (Jan of year i)
	xp = zsim[1] # Previous value initialization
	
# Monthly flow simulations
	
		for (j in 2:N) { 
	
		mo = j%%12 # month we are simulating
		imon=ifelse(mo==0,12,mo)
		
		if (mo != 1){ # if mo = 0 (Dec) 
		data = flow[,1:12]  # data is the same as test (Jan - Dec)
		}	else { # Jan
		data = cbind(flow[2:nyrs,1],flow[1:nyrs1,2:12])
		}
		# if mo = 1 (JAn), 1st col of data will be March, last col will be Feb
		mo1 = mo-1
		if(mo == 0)mo1=11
		if(mo == 1)mo1=12
		xdist = order(abs(xp - data[,mo1]))
		
#	if (mo == 1) {
#	xdist = order(abs(xp - data[,mo1]))}
#	if (mo >1) {
#	xx=rbind(xp, data[,(12-ilag):11]) 
#	xdist = order(as.matrix(dist(xx))[1,2:nyrs+1]) }

		xx=runif(1,0,1) # generate random number
		xy=c(xx,W)
		xx=rank(xy) # get rank with respect to W
		i1=xdist[xx[1]] # distance order at given rank
	
		zsim[j]=data[i1,imon] # select data at given distance
		xp = zsim[j] # store previous year for next it.
	} # end j loop

	simdismon = matrix(zsim, ncol=12, byrow = TRUE) # makes a 12 column matrix with jan thru dec
	maysim = simdismon[,5]
	simpdf[isim,]=sm.density(maysim,eval.points=xeval,display="none")$estimate

	# Fill statistics for each month
  for(j in 1:12){
    armean[isim,j]=mean(simdismon[,j])
    armax[isim,j]=max(simdismon[,j])
    armin[isim,j]=min(simdismon[,j])
    arstdev[isim,j]=sd(simdismon[,j])
    arskw[isim,j]=skewness(simdismon[,j])
  }
  arcor[isim,1]=cor(simdismon[-nyrs,12],simdismon[2:nyrs,1]) #cor dec-jan
  for(j in 2:12){ # rest of pairs
    j1=j-1
    arcor[isim,j]=cor(simdismon[,j],simdismon[,j1])
  }
  
# Anual flow simulations
  
  y.sim = 1:nyrs # n of anual flow time series
  y.sim[1] = flow[i,"year"]  # initial flow (Year i)
  y.p = y.sim[1] # Previous value initialization
  y.data=flow$year # dataset is fixed for anual ts.
  
  for(j in 2:nyrs){
    y.dist = order(abs(y.p - y.data))
    i1 = runif(1,0,1) %>% c(.,W) %>% rank(.) %>% {y.dist[.[1]]}
    y.sim[j]=y.data[i1] # select y.data at given distance
    y.p = y.sim[j] # store previous year for next it.
    
  }
  
  # Get anual PDF
  year.pdf[isim,]=sm.density(y.sim,eval.points=
                            yeval,display="none")$estimate
  
  # Calculate statistics
  ar.year.stat[isim,"mean"]=mean(y.sim)
  ar.year.stat[isim,"max"]=max(y.sim)
  ar.year.stat[isim,"min"]=min(y.sim)
  ar.year.stat[isim,"stdev"]=sd(y.sim)
  ar.year.stat[isim,"skew"]=skewness(y.sim)
  ar.year.stat[isim,"cor"]=cor(y.sim[-nyrs],y.sim[2:nyrs])

} # end isim loop

```

We add the statistics from historical data in the first row of the tables

```{r }
# Compute statistics from the historical data.
obsmean=1:12
obsstdev=1:12
obscor=1:12
obsskw=1:12
obsmax=1:12
obsmin=1:12

for(i in 1:12){
  obsmax[i]=max(flow[,i])
  obsmin[i]=min(flow[,i])
  obsmean[i]=mean(flow[,i])
  obsstdev[i]=sd(flow[,i])
	obsskw[i]=skewness(flow[,i])
}

obscor[1]= cor(flow[-nyrs,12], flow[2:nyrs,1])
for(i in 2:12){
  i1=i-1
  obscor[i]=cor(flow[,i], flow[,i1])
}

# bind the stats of the historic data at the top..
armean=rbind(obsmean,armean)
arstdev=rbind(obsstdev,arstdev)
arskw=rbind(obsskw,arskw)
arcor=rbind(obscor,arcor)
armax=rbind(obsmax,armax)
armin=rbind(obsmin,armin)

# anual flow binding
year.stat=c(mean(flow$year),sd(flow$year),min(flow$year),
            max(flow$year),skewness(flow$year),
            cor(flow$year[-nyrs],flow$year[2:nyrs]))

ar.year.stat = rbind(year.stat,ar.year.stat)

```

## Plot statistics from simulations
* **Create boxplots of annual and monthly, mean, variance, skew, lag-1 correlation, minimum, maximum and PDFs of May and annual flows. Comment on what you observe and also on why some of the monthly statistics are not captured.**

```{r}
# function to plot boxplots with the structure: hist. in first row
plot.bp = function(matrix,name){
  xmeans=as.matrix(matrix)
  n=length(xmeans[,1])
  xmeans1=as.matrix(xmeans[2:n,]) #the first row is the original data
  xs=1:12
  zz=boxplot(split(xmeans1,col(xmeans1)), plot=F, cex=1.0)
  zz$names=rep("",length(zz$names))
  z1=bxp(zz,ylim=range(xmeans),xlab="",ylab="",cex=1.00)
  points(z1,xmeans[1,],pch=16, col="red")
  lines(z1,xmeans[1,],pch=16, col="gray")
  title(main=name)
}
```

The plots for the statistics of the simulated time series (shown as boxplots) vs. the historical data (shown as points and lines) are reproduced below:

```{r fig.height=8}
par(mfrow=c(3,2))
plot.bp(armean,"Mean")
plot.bp(arstdev,"Standard Deviation")
plot.bp(armin,"Min")
plot.bp(armax,"Max")
plot.bp(arskw,"Skews")
plot.bp(arcor,"Lag-1 correlation")
```

All the statistics are well fitted with respect the historical data.

The anual statistics are similarly represented below:

```{r}
par(mfrow=c(2,3))
plot.bp(ar.year.stat[,"mean"],"Mean")
plot.bp(ar.year.stat[,"stdev"],"Standard Deviation")
plot.bp(ar.year.stat[,"min"],"Min")
plot.bp(ar.year.stat[,"max"],"Max")
plot.bp(ar.year.stat[,"skew"],"Skews")
plot.bp(ar.year.stat[,"cor"],"Lag-1 correlation")

```

For the anual flow, this is no longer true. Some statistics are not poorly fitted.

```{r}
xdensityorig = flow$may %>% sm.density(.,eval.points=xeval,display="none") %>%
  .$estimate

plot.pdf = function(eval,histPDF,simPDF){
  xeval = eval
  plot(xeval,histPDF,pch=".",col="red",ylim=range(simPDF,histPDF),
       xlab="",ylab = "")
  for(i in 1:nsim)lines(xeval,simPDF[i,],col='lightgrey',lty=3)
  lines(xeval,histPDF,lwd=3,col="red")
  title(main="Historical vs. simulated PDF")
}
```
```{r}
plot.pdf(xeval,xdensityorig,simpdf)
```

The simulated PDF shows the bimodal behavior, and does not respond to a probabilistic distribution, as it happened to parametric AR.

```{r}
year.density = flow$year %>% sm.density(.,eval.points=yeval,
               display="none") %>% .$estimate
```
```{r}
plot.pdf(yeval,year.density,year.pdf)
```

Many results are repeated (only 111 values for 100 simulations of 111 time series), creating a crisp PDF. Not realistic.

## Advantages/disadvantages with a nonparametric approach. 

When the dataset is large enough, the fitted PDF and statistics over perform the parametric modelling.

When the opposite occurs, many values are repeatedly picked, resulting in biased PDF and statistics.



