---
title: "03 Multivarite Simulation Copula"
output: html_document
#theme: flatly
---

# Multivariate Simulation - Copulas

**Another approach to simulating the monthly streamflow is using Copulas.**
**Fit appropriate marginal PDFs for each monthly streamflow**

### Data import and setup

```{r results=FALSE}
# Load libraries
libr=c("magrittr","sm","MASS","ks","copula","moments")
options(warn=1)
suppressPackageStartupMessages(lapply(libr, require, character.only = TRUE))
```

```{r}
# import and set up flow data
flow = read.table(
  "http://civil.colorado.edu/~balajir/CVEN6833/HWs/HW-3-2018/LeesFerry-monflows-1906-2016.txt")

flow = flow[,2:13] %>% `rownames<-`(flow[,1]) %>%
  setNames(.,c("jan","feb","mar","apr","may","jun",
                      "jul","aug","sep","oct","nov","dec")) %>%
                      {./10^6} # convert AF to MAF
head(flow,n=1L) # show values
tail(flow,n=1L)
```

## Fit a Copula to the joint CDF (copula by pairs)

As a first approach, we will use copula by pairwise comparison (lag-1 model).

```{r}
# Function to fit copula estimator for a pair of time series
fit2copula = function(data1,data2){
  u = pobs(cbind(data1,data2))
  fc=normalCopula(dim=2,disp='un')
  fnc = fitCopula(fc,u)
  return(fnc)
  }

# Function to generate simulated values from fitted copula 
sim.copula = function(fnc,data2,N){
  cop.sim = rCopula(N,normalCopula(fnc,dim=2,dispstr='un'))
  data.sim = quantile(data2,cop.sim[,2])
  # quan.sim = qkde(cop.sim[,2],fhat)
  return(data.sim) # return(quan.sim)
}
```
 
## Simulate from the Copula and invert
 
```{r}
# Simulation paramenters
nsim=250
N=length(flow[,1]) # years

armean=matrix(0,nsim,12)	#matrices to store the statistics
arstdev=matrix(0,nsim,12)
arcor=matrix(0,nsim,12)
arskw=matrix(0,nsim,12)
armax=matrix(0,nsim,12)
armin=matrix(0,nsim,12)

xeval = matrix(0,nrow = 100,ncol = 12)  # Points where the PDF is evaluated 

for(i in 1:12){
  xeval[,i]=seq(min(flow[,i])-0.25*sd(flow[,i]),
          max(flow[,i])+0.25*sd(flow[,i]),length=100)
}
simpdf = array(0,c(100,12,nsim)) # Array to store simulated PDF
xsim = array(0,c(N,12,nsim)) # Array to store simulated val

# Simulation initialization parameters

ncop = 12 # generate pairs of data to use in copula
data1 = data2 = array(NA,c(N,1,ncop)) # data 1 / 2: prev / current month
fnc = 1:ncop # fitted normal copulas for every pair

for(icop in 1:ncop){
  if(icop==1){
    data1[-N,1,icop] = flow$dec[-N] # dec of previous year
    data2[-N,1,icop] = flow$jan[2:N] # jan of next year
    fnc[icop] = fit2copula(data1[-N,,icop],data2[-N,,icop])@estimate
  }else{
    data1[,1,icop] = flow[,icop-1] # as given
    data2[,1,icop] = flow[,icop] # as given
    fnc[icop] = fit2copula(data1[,,icop],data2[,,icop])@estimate
  }
}

for(isim in 1:nsim){ # simulation loop
  
  for(j in 1:12){ # get each month as copula with previous month
    if(j==1){ # dec-jan copula
      xsim[,j,isim]=sim.copula(fnc[j],data2[-N,,j],N) # remove NA year 111
    }else{
      xsim[,j,isim]=sim.copula(fnc[j],data2[,,j],N)   # simulated values
    }
    simpdf[,j,isim]=sm.density(xsim[,j,isim],        # simulated PDF
                                 eval.points=xeval[,j],display="none")$estimate
    # fill statistics
    armean[isim,j]=mean(xsim[,j,isim])   
    armax[isim,j]=max(xsim[,j,isim])
    armin[isim,j]=min(xsim[,j,isim])
    arstdev[isim,j]=sd(xsim[,j,isim])
    arskw[isim,j]=skewness(xsim[,j,isim])
    if(j>2)arcor[isim,j]=cor(xsim[,j,isim],xsim[,j-1,isim])
  }
  arcor[isim,1]=cor(xsim[-N,12,isim],xsim[2:N,1,isim])
  # print(isim)
}
```

We now add the statistics from the original dataset and bind it with the simulation st.

```{r}
# Compute statistics from the historical data.
obsmean=1:12
obsstdev=1:12
obscor=1:12
obsskw=1:12
obsmax=1:12
obsmin=1:12

for(i in 1:12){
  obsmax[i]=max(flow[,i])
  obsmin[i]=min(flow[,i])
  obsmean[i]=mean(flow[,i])
  obsstdev[i]=sd(flow[,i])
	obsskw[i]=skewness(flow[,i])
}

obscor[1]= cor(flow[-N,12], flow[2:N,1])
for(i in 2:12){
  obscor[i]=cor(flow[,i], flow[,i-1])
}

# bind the stats of the historic data at the top..
armean=rbind(obsmean,armean)
arstdev=rbind(obsstdev,arstdev)
arskw=rbind(obsskw,arskw)
arcor=rbind(obscor,arcor)
armax=rbind(obsmax,armax)
armin=rbind(obsmin,armin)
```

## Boxplot the statistics listed in problem 1 and compare with the results from the previous methods

We repeat the code from problem 1 to do so.

```{r}
# function to plot boxplots with the structure: hist. in first row
plot.bp = function(matrix,name){
  xmeans=as.matrix(matrix)
  n=length(xmeans[,1])
  xmeans1=as.matrix(xmeans[2:n,]) #the first row is the original data
  xs=1:12
  zz=boxplot(split(xmeans1,col(xmeans1)), plot=F, cex=1.0)
  zz$names=rep("",length(zz$names))
  z1=bxp(zz,ylim=range(xmeans),xlab="",ylab="",cex=1.00)
  points(z1,xmeans[1,],pch=16, col="red")
  lines(z1,xmeans[1,],pch=16, col="gray")
  title(main=name)
}
```

```{r fig.height=8}
par(mfrow=c(3,2))
plot.bp(armean,"Mean")
plot.bp(arstdev,"Standard Deviation")
plot.bp(armin,"Min")
plot.bp(armax,"Max")
plot.bp(arskw,"Skews")
plot.bp(arcor,"Lag-1 correlation")
```

As observed, mean, st.dev, max and skews are well fit. Min values are overestimated, and lag-1 is not captured. In general, the statics are better than the ones using autoregressive models.

```{r}
xdensityorig = matrix (0,nrow = 100,ncol = 12) #initialize original PDF

for(j in 1:12){ # obtain original PDF
  xdensityorig[,j] = flow[,j] %>% sm.density(.,eval.points=xeval[,j],display="none") %>%
  .$estimate
}

plot.pdf = function(eval,histPDF,simPDF,title){
  xeval = eval
  plot(xeval,histPDF,pch=".",col="red",ylim=range(simPDF,histPDF),
       xlab="",ylab = "")
  for(i in 1:nsim)lines(xeval,simPDF[,i],col='lightgrey',lty=3)
  lines(xeval,histPDF,lwd=3,col="red")
  title(main=title)
}
```
```{r }
par(mfrow=c(4,3))
par(mar=c(2,2,2,2))

for(j in 1:12){
  plot.pdf(xeval[,j],xdensityorig[,j],simpdf[,j,],colnames(flow[j]))
}
  
```

The simulated PDF do not respond to a probabilistic distribution, sitting closely around the empirical PDF.  

The result is more precise that the one using autoregressive models.

## 12 months copula

Now we will use the full copula between all months.

```{r}
u = pobs(flow) # convert to pseudo-observations
fc12=normalCopula(dim=12,disp='un') # normal copula (66)
fnc12 = fitCopula(fc12,u) # estimate from copula

dat_sim = array(0,c(N,12,nsim))
cop_pdf = array(0,c(100,12,nsim))

carmean=matrix(0,nsim,12)	#matrices to store the statistics
carstdev=matrix(0,nsim,12)
carcor=matrix(0,nsim,12)
carskw=matrix(0,nsim,12)
carmax=matrix(0,nsim,12)
carmin=matrix(0,nsim,12)

for(isim in 1:nsim){
  cop_sim = rCopula(N,normalCopula(fnc12@estimate,dim=12,dispstr='un'))

  for(j in 1:12){
    dat_sim[,j,isim] =  quantile(flow[,j],cop_sim[,j]) 
    cop_pdf[,j,isim]=sm.density(dat_sim[,j,isim],        # simulated PDF
                                 eval.points=xeval[,j],display="none")$estimate
    # fill statistics
    carmean[isim,j]=mean(dat_sim[,j,isim])   
    carmax[isim,j]=max(dat_sim[,j,isim])
    carmin[isim,j]=min(dat_sim[,j,isim])
    carstdev[isim,j]=sd(dat_sim[,j,isim])
    carskw[isim,j]=skewness(dat_sim[,j,isim])
    if(j>2)carcor[isim,j]=cor(dat_sim[,j,isim],dat_sim[,j-1,isim])
  }
  carcor[isim,1]=cor(dat_sim[-N,12,isim],dat_sim[2:N,1,isim])
  # print(isim)
}

```

```{r}
# bind the stats of the historic data at the top..
carmean=rbind(obsmean,carmean)
carstdev=rbind(obsstdev,carstdev)
carskw=rbind(obsskw,carskw)
carcor=rbind(obscor,carcor)
carmax=rbind(obsmax,carmax)
carmin=rbind(obsmin,carmin)
```

```{r fig.height=8}
par(mfrow=c(3,2))
plot.bp(carmean,"Mean")
plot.bp(carstdev,"Standard Deviation")
plot.bp(carmin,"Min")
plot.bp(carmax,"Max")
plot.bp(carskw,"Skews")
plot.bp(carcor,"Lag-1 correlation")
```

The results are at least as good as the copula by pairs. Also, the lag-1 correlation is captured (with the exception of jan and feb).

```{r fig.height=8}
par(mfrow=c(4,3))
par(mar=c(2,2,2,2))

for(j in 1:12){
  plot.pdf(xeval[,j],xdensityorig[,j],cop_pdf[,j,],colnames(flow[j]))
}
  
```

The simulated PDF are also really close to the empirical PDF, being able to mimic the multimodal behavior of February, May, July, etc.

 

