---
title: "10-HMM-GEV"
output: html_document
---

# HMM with GEV component distributions

**Simulate the annual maximum flow on Clark Fork River with a Hidden Markov Model (Markov Chain + Resampling) using GEV component distributions.**

The _HiddenMarkov::dthmm_ function allows to use custom distributions for the expectation maximization algorithm via _HiddenMarkov::Mstep_ function. The function is called in the _distn_ argument of _dthmm_.

```{r results = FALSE}

# Load libraries, data

set.seed(1) # allows repetition of random generations
# libraries
libr=c("HiddenMarkov","extRemes")
options(warn=-999)
suppressPackageStartupMessages(lapply(libr, 
                require, character.only = TRUE))

data=read.table(
  "http://civil.colorado.edu/~balajir/CVEN6833/HWs/HW-3-2018/Cfork-enso-pdo.txt")
colnames(data) = c("year","CFR","ENSO","PDO")
rownames(data) = data$year
data = data[,!colnames(data) %in% "year"]
```

```{r eval=FALSE}
# Try a dthmm with exp distr

family <- "exp"  # underlying distribution for hmm
discrete <- FALSE
aic1=c()
x=data$CFR+abs(min(data$CFR))
for(imodel in 2:6){
  m <- imodel      #model order to fit
  stationary <- F   # use a stationary distribution of mixtures
      # different initial condition types when family == "norm"
  ic <- "same.sd"#c("same.sd","same.both","both.diff")
  fd.name <- ifelse(family == "exp", "exponential", family)
  Pi <- Pi_init(m)     # T.P.M.
  delta <- delta_init(m)
  ######
  # PROBLEM HERE#
  #####
  pars <- get.named.parlist(x, m, fd.name, lower=.0, ic) # This needs to be GEV
  #,start=list(shape1=2,shape2=2))
  # set up the model
  hmm <- dthmm(x, Pi=Pi, delta=delta, family, pars, nonstat=!stationary, discrete = discrete)
  sink("p.4.hmm.fit")
  if(imodel < 2){
   hmm <- BaumWelch(hmm, bwcontrol(maxiter = 1000,
          posdiff=TRUE,converge = expression(diff > tol)))
   } else {
   hmm <- BaumWelch(hmm, bwcontrol(maxiter = 1000, tol = 1e-08))
   }
  sink()
  # get the hidden states from the fitted model 
  # Global decoding. To get the probability of being in a state: hmm$u
  decoding <- Viterbi(hmm)
  # get AIC
  aic <- AIC(hmm)
  aic1=c(aic1,aic)
}

```

```{r P.10.functions}
# source functions
# takend from "http://civil.colorado.edu/~balajir/CVEN6833/R-sessions/session3/files-4HW3/lib.R"

Pi_init <- function(n,type='uniform'){
    matrix(rep(1/n,n^2),n)}

delta_init <- function(n, type='uniform'){
    d <- rnorm(n)^2
    d/sum(d)}

ntile.ts <- 
function(x, n, limit.type = 'prob', tie = 1, altobs = NULL ){
    # returns an integer vector corresponding to n states broken by equal 
    # probability or equal distance
    #
    limit <- 
    if(limit.type == 'prob') 
      quantile(x,seq(0,1,1/n))
    else if(limit.type == 'equal')
      seq(min(x),max(x),by=diff(range(x))/n)

    if(!is.null(altobs)) limit <- quantile(altobs,seq(0,1,1/n))
    
    b <- integer(length(x))
    
    for(i in 1:(n+1)){
        filter <- 
        if(tie == 1) 
            x >= limit[i] & x <= limit[i+1]
        else 
            x > limit[i] & x <= limit[i+1]
    
        #only need to set the 1's because b is already 0's
        b[filter] <- as.integer(i-1)
    }
    
    if(class(x) == 'ts') 
        return(ts(b,start=start(x),end=end(x))) 
    else 
        return(b)
}

get.named.parlist <- function(x,m,dist,ic,...){
    require(MASS)
    fit <- fitdistr(x,dist,...)
    np <- length(fit$estimate)
    pars <- vector('list',np)
    names(pars) <- names(fit$estimate)

    init <- lapply(fit$estimate,max)
    names(init) <- names(fit$estimate)

    for(j in 1:m){
        #print(j)
        #browser()

        #browser()
        this.fit <- fitdistr(x[ntile.ts(x,m) == (j-1)],dist,init,...)
        #for(k in 1:np)
        #    pars[[k]][j] <- this.fit$estimate[k]
        for(k in 1:np)
            pars[[k]][j] <- fit$estimate[k]
        if(dist == 'normal'){
            if(ic == 'same.both'){
                pars[[k]][j] <- mean(x)
                pars[[k]][j] <- sd(x)
            } else if( ic == 'same.sd'){
                pars[[k]][j] <- mean(x[ntile.ts(x,m) == (j-1)])
                pars[[k]][j] <- sd(x)
            }else{
                pars[[k]][j] <- mean(x[ntile.ts(x,m) == (j-1)])
                pars[[k]][j] <- sd(x[ntile.ts(x,m) == (j-1)])
            }
        }
    }
    pars
}

AIC.dthmm <- function(x){
  ## Return the Akaieke Information criterion value for a fitted discrete 
  ## time hidden markov model from the HiddenMarkov package
  
    # Model order
  m <- length(x$delta)
    # Log Liklihood value
  LL <- x$LL
    # number of parameters
  p <- m+m^2
    # AIC
  -2*LL + 2*p
}

ggplot_stationary_hmm <- function(x,binwidth=NULL,res=1000,cols=NULL,...){
    
    m <- length(x$delta)
    dens <- matrix(0,nrow=m+1,ncol=res)
    r <- extendrange(x$x,f=.05)
    xrange <- seq(r[1],r[2],len=res)
    delta <- statdist(x$Pi)
    if(is.null(binwidth)) binwidth <- diff(range(x$x))/8
    for(i in 1:m){
        
        if(x$distn == 'gamma'){
            dens[i,] <- delta[i]*dgamma(xrange,shape=x$pm$shape[i],rate=x$pm$rate[i])
        }else if(x$distn == 'norm'){
            dens[i,] <- delta[i]*dnorm(xrange,mean=x$pm$mean[i],sd=x$pm$sd[i])
        }else{
            stop('Distribution not supported')
        }
            
        dens[m+1,] <- dens[m+1,] + dens[i,]
    }

    p <- ggplot()+
        geom_histogram(data=data.frame(x=as.vector(x$x)),aes(x=x,y=..density..),
            binwidth=binwidth,fill='white',color='black')+
        theme_bw()

    dt <- data.table(x=numeric(0),y=numeric(0), state=integer(0))
    for(i in 1:m)
        dt <- rbind(dt, data.table(x=xrange,y=dens[i,], state=i))
    dt$state <- factor(dt$state)
    
    p <- p + geom_line(data=dt,aes(x=x,y=y,color=state)) + 
        geom_line(data=data.frame(x=xrange,y=dens[m+1,]),aes(x=x,y=y),color='black',size=1) + 
        scale_color_tableau() +
        scale_x_continuous(limits=r)
    p
    
}

statdist <- function(tpm){
    
    m <- nrow(tpm)
    ones <- rbind(rep(1,m))
    I <- diag(rep(1,m))
    U <- matrix(rep(1,m^2),m)
    as.vector(ones %*% solve(I - tpm + U))
    
}
```